---
title: "Supplementary Note 1: R code to compare lme4qtl and pedigreemm R packages "
author: "Andrey Ziyatdinov"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: false
  html_document:
    fig_caption: yes
    toc: true
    number_sections: false
bibliography: ref.bib    
---

```{r options, echo = F}
opts_chunk$set(comment = NA, results = 'markup', tidy = F, message = F, warning = F, echo = T,
  fig.width = 3, fig.height = 3, cache = T)
```

# About

The R package `pedigreemm` was first in extending the `lme4` R package
for particular applications in the animal breeding field [@Vazquez2010].
Custom covariance (genetic additive) matrix 
are defined using the pedigree annotation information
(`pedigree` argument of `pedigreemm` function).
Although the `lme4qtl` package borrows the same idea of `pedigreemm`,
`lme4qtl` provides a larger list of genetic models
that are *not* possible with `pedigreemm`.
In particular, these models include:

- models with a single or several custom covariances 
  (not necessary linked to pedigree information);
- models with random slopes and other similar models 
  like gene-by-environment interaction models;
    - the restriction on model parameters, 
      e.g. the correlation coefficient is zero,
      is supported.

Here, we show models that are available with `lme4qtl` and not with `pedigreemm`.

# Include

First, we load R packages necessary for data analysis.

```{r inc0}
library(Matrix)
library(magrittr)
```

```{r inc}
library(pedigreemm)

library(lme4qtl)
```

# Data

We use an example data set `milk` from the `pedigreemm` package.
See `?milk` for description.

Here, we work on a subset of this dataset (`milk_subset`)
to reduce the computation time.

```{r milk}
data(milk)

milk <- within(milk, {
  id <- as.character(id)
  sdMilk <- milk / sd(milk)
})

ids <- with(milk, id[sire %in% 1:3]) # for more ids: c(1:3, 319-321)
milk_subset <- subset(milk, id %in% ids)

milk_subset <- within(milk_subset, {
   herd <- droplevels(herd)
   herd_id <- paste0("herd", seq(1, length(herd)))
})
```

## Covariance matrices

A mixed model we are going to fit will have two random effects,
groupings based on two ID variables:

- `id`, a numeric identifier of cow (the genetic additive effect);
- `herd`, a factor indicating the herd (the shared environmental effect).

Further we derive the covariance matrices (among samples)
due to these two random effects.

```{r A_herd}
A_herd <- with(milk_subset, model.matrix(~ herd - 1)) %>% 
  tcrossprod %>% Matrix

rownames(A_herd) <- milk_subset$herd_id
colnames(A_herd) <- milk_subset$herd_id
```

```{r A_gen}
A_gen <- getA(pedCowsR)

stopifnot(all(ids %in% rownames(A_gen)))
ind <- rownames(A_gen) %in% ids

A_gen <- A_gen[ind, ind]
```


```{r A_mat, fig.show = "hold", out.width = "45%"}
image(A_herd, main = "A_herd")
image(A_gen, main = "A_gen")
```


# Models

## A single kinship matrix

Both packages can fit a basic model with a single genetic effect,
for which the `pedigreemm` R package was sought.

```{r m1_pmm}
m1_pmm <- pedigreemm(sdMilk ~ lact + log(dim) + (1|id) + (1|herd),
  data = milk_subset, pedigree = list(id = pedCowsR))

VarCorr(m1_pmm)
```   

```{r m1_relmat}
m1_relmat <- relmatLmer(sdMilk ~ lact + log(dim) + (1|id) + (1|herd),
  data = milk_subset, relmat = list(id = A_gen))

VarCorr(m1_relmat)
``` 

We see that the estimation of variance components from both packages are identical.

## A single custom covariance matrix

`lme4qtl` packages allows for custom covariance matrices, 
while `pedigreemm` does not.

```{r m2}
m2_lmer <- lmer(sdMilk ~ (1|herd), milk_subset)
VarCorr(m2_lmer)

m2_relmat <- relmatLmer(sdMilk ~ (1|herd_id), milk_subset, 
  relmat = list(herd_id = A_herd))
VarCorr(m2_relmat)

(try(m2_pmm <- pedigreemm(sdMilk ~ (1|herd_id), milk_subset, 
  pedigree = list(herd_id = A_herd))))
```

```{r Ztlist_34, fig.show = "hold", out.width = "45%"}
getME(m2_lmer, "Ztlist")[[1]] %>% crossprod %>% image
getME(m2_relmat, "Ztlist")[[1]] %>% crossprod %>% image
```

### Rank deficiency

`A_herd` is a low-rank matrix, but `lme4qtl` is able to deal with this rank deficiency situation
by replacing the Cholesky decomposition by the EVD operation.
The `pedigreemm` package only uses the Cholesky decomposition.

```{r rank}
A_herd %>% dim

A_herd %>% as.matrix %>% qr %$% rank
```

## A single kinship matrix + random slope

Complex models are possible with `lme4qtl`, for example, those with a random slope effect.

```{r m3_rs}
m3_relmat <- relmatLmer(sdMilk ~ lact + log(dim) + (1 + lact|id) + (1|herd),
  data = subset(milk_subset, relmat = list(id = A_gen)))
VarCorr(m3_relmat)

(try(m3_pmm <- pedigreemm(sdMilk ~ lact + log(dim) + (1 + lact|id) + (1|herd),
  data = milk_subset, pedigree = list(id = pedCowsR))))
```   

### Restriction on model parameters

```{r m3_r}
m3_relmat_rho0 <- relmatLmer(sdMilk ~ lact + log(dim) + (1 + lact|id) + (1|herd),
  data = subset(milk_subset, relmat = list(id = A_gen)),
  vcControl = list(rho0 = list(id = 2)))
VarCorr(m3_relmat_rho0)
```


## Two covariance matrices

```{r m56}
m5 <- relmatLmer(sdMilk ~ (1|herd) + (1|id), milk_subset, 
  relmat = list(id = A_gen))
VarCorr(m5)

m6 <- relmatLmer(sdMilk ~ (1|herd_id) + (1|id), milk_subset, 
  relmat = list(herd_id = A_herd, id = A_gen))
VarCorr(m6)
```


```{r Ztlist_56, fig.show = "hold", out.width = "45%"}
getME(m5, "Ztlist")[[1]] %>% crossprod %>% image
getME(m6, "Ztlist")[[2]] %>% crossprod %>% image
```

```{r Ztlist_56_2, fig.show = "hold", out.width = "45%"}
getME(m5, "Ztlist")[[2]] %>% crossprod %>% image
getME(m6, "Ztlist")[[1]] %>% crossprod %>% image
```



```{r m3}
(try(m3 <- pedigreemm(sdMilk ~ lact + log(dim) + (1|id) + (1|herd_id),
  data = milk_subset, pedigree = list(id = pedCowsR, herd_id = A_herd))))
``` 

# References


